Polynomial Chaos for the Estimation of Manning's Friction

### 2013-12-19 paper/analysis.tex
last changes from my side... i think

\subsubsection{Statistical analysis}
The mean of the QoI and its standard deviation is computed
from the PC coefficients as indicated in Equation~\eqref{eq:mean} and Equation~\eqref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the calculated PC mean water surface elevation along with two standard deviations
bounds at the four gauges as indicated in each panel.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
reflected in a significant standard deviation as seen earlier from the realizations shown 
in Figure~\ref{fig:rlzs}. An interesting observations is that the
standard deviation in water surface elevations vanishes at few times instances
during the tsunami. This can be explained as ......

>>>>

\subsubsection{Statistical analysis}
The mean of the QoI and its standard deviation is computed
from the PC coefficients as indicated in Equation~\eqref{eq:mean} and Equation~\eqref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the calculated PC mean water surface elevation along with two standard deviations
bounds at the four gauges as indicated in each panel.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
reflected in a significant standard deviation as seen earlier from the realizations shown 
in Figure~\ref{fig:rlzs}. An interesting observations is that the
standard deviation in water surface elevations vanishes at few times instances
during the tsunami. \alert{This can be explained as ...... }

These statistics can be used to compare with the 
(DART) buoys observations. Figure~\ref{fig:compare} 
the PC mean water surface elevation with the observed 
water surface elevation at the different gauges locations
with error bars indicated at few time instances. 
The plots show a good agreement between the PC simulations and the 
observations at different locations and at different times.

### 2013-12-19 paper/mcmc.tex
last changes from my side... i think

Before attempting solving the inverse problem,
we need to make sure that the observed data are
comparable to the simulated one. To this end, Figure~\ref{fig:compare}
compares PC mean water surface elevation profiles (with error bars)
with their counterparts from the observed gauge locations.
The plots show a good agreement between the simulations and the 
observations at different locations and at different times. 
The same comparison can be recast as a scatter plot shown in 
Figure~\ref{fig:scatter}. The difference between the observations
and simulations is attributed to the uncertainty in the 
Manning's N coefficient and to the measurement errors to be quantified
in this section.

The surrogate model is exploited in the Bayesian inference of the Manning's 
roughness coefficient.  To this end, an adaptive MCMC method is used to sample 
the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
update the Manning's roughness coefficient distributions in light of the 
observed data. This sampling, demanding tens of thousands of forward simulations, 
would have been prohibitively expensive in the
absence of the surrogate, as the generation of each sample would have required an
independent \geoclaw realization. The surrogate provides a computationally
efficient alternative, and requires only evaluating the PCE for different values 
of the seed $\xxi$. The setup of the Bayesian inference problem,
MCMC chains for the input drag parameters and the corresponding posterior distributions are 
presented and discussed in this section.

The observed data collected at different locations 
in the likelihood function (Equation~\eqref{eq:likelihood}) to update the input parameters.
MCMCs of $50,000$ iterations are obtained for the Manning roughness coefficients: 
$N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
shows the sample chains for
the input parameters for different iterates of the MCMC algorithms. 
The different panels
shows well-mixed chains for all parameters.
However, all chains span the entire range
of the prior, and so it appears that the observations are not informative 
concerning this uncertain input.  $N_{2}$ chain appears to be concentrated in the lower end of the
parameter range, with values between 0.005 and 0.1.
The chain for the variance is also shown in 
Figure~\ref{fig:mcmc} and appears to be well mixed.

The computed MCMC chains can be readily used to determine the posterior 
distribution; kernel density estimation (KDE) is used for this purpose
~\citep{Parzen1962,Silverman1986}(The first 5000 iterates, associated 
with the burn-in period, are discarded). The resulting posterior pdfs 
of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
in Figure~\ref{fig:pdfs}.  As expected from the chains shown in Figure
~\ref{fig:mcmc}, the posterior pdf of $N_2$, exhibits well-defined peak, 
with a Maximum A Posteriori (MAP) estimate of around 0.011; an extended tail 
towards the higher Manning's roughness coefficient values is also observed.
In contrast, the posterior pdfs of $N_1$ appear to be fairly flat, 
and similar to the uniform prior. This is an indication that 
the observed data were not useful to refine our prior knowledge for $N_1$.
While for $N_3$, we observed a posterior that has a well defined peak
of 0.185 but no clear pdf shape.
The posterior distribution of the variance is also shown in Figure~\ref{fig:mcmc}. 
Taking the square root of the MAP values yields the water surface elevation standard 
deviation, and the latter is a reflection of the mismatch between the model and 
observed data estimated to be 0.145~$m$.

>>>>

Before we attempt solving the inverse problem of parameter estimation, 
we first check the ability of \geoclaw to simulate water surface elevation
realistically. To this end, we present a comparison between the 
(DART) buoys observations  and their \geoclaw model counterparts
for a reference simulation in which the  the Manning's N coefficients were set to their optimal values $N_1=N_2=N_3=0.015$. Figure~\ref{fig:scatter} 
shows a scatter plot that compares the observed 
water surface elevation at the different gauges locations and the  \geoclaw model counterparts. 
The plots show a good agreement between the simulations and the 
observations at different locations and at different times. 
The difference between the observations and simulations is attributed to the uncertainty in the 
Manning's N coefficient and to the measurement errors.
\alert{An interesting observation is the more scatter in the data collection by gauge 21418 which is the closest to
the epicenter of the earthquake and therefore receiving more impact by the tsunami.}

Bayesian inference is now used to estimate the Manning's 
roughness coefficient ranges that minimize this scatter between 
observed and modeled water surface elevation, and to estimate the variance of the noise in the measured data.
To this end, an adaptive MCMC method is used to sample 
the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
update the Manning's roughness coefficient distributions in light of the 
observed data. This sampling requires tens of thousands of 
\geoclaw model runs that are prohibitively expensive as each MCMC 
sample requires an independent \geoclaw realization. Instead,
the surrogate model created using PC expansions provides a computationally
efficient alternative that requires only evaluating the PC expansion
for different values of the germ $\xxi$.

The MCMC algorithm required $50,000$ iterations to guarantee convergence
and to obtain the posteriors of the Manning roughness coefficients: 
$N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
shows the sample chains for the input parameters for different iterations of the MCMC algorithms. 
The different panels shows well-mixed chains for all input parameters as well as the variance $\sigma^2$.
The chain of  $N_{1}$ span the entire range of the prior meaning the observations are not informative 
concerning this uncertain input.  On the contrary, the $N_{2}$ chain appears to be concentrated in the 
lower end of the parameter range, with values between 0.005 and 0.1. 
While for $N_{3}$, the chain appears to be concentrated in the 
upper end of the parameter range. Finally, the chain for the variance 
appears to be well mixed with a well defined posterior range between 0.02 and 0.022.

Next, the computed MCMC chains are used to determine the posterior 
distributions using kernel density estimation (KDE)
~\citep{Parzen1962,Silverman1986}.  The resulting posterior pdfs 
of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
Note that the first 5000 iterates, associated with the burn-in period, were discarded.  
As expected from the chains shown in Figure
~\ref{fig:mcmc}, the posterior pdfs of $N_1$ appear to be fairly flat, 
and similar to the uniform prior; an indication that 
the observed data were not useful to refine our prior knowledge for $N_1$.,
In contrast, the posterior pdf of $N_2$ exhibits well-defined peak, 
with a Maximum A Posteriori (MAP) estimate of around 0.011
and an extended tail towards the higher Manning's roughness coefficient values.
While for $N_3$, we observed a posterior that has a well defined peak
of 0.185 but no clear pdf shape. The posterior distribution of the variance 
appears to be well-defined of Gaussian-like shape. 
The MAP value of $\sigma^2=0.021$ can be used to estimate the water surface elevation standard 
deviation estimated to be 0.145~$m$. 
This value is a reflection of the mismatch between the model and 
observed data.

### 2013-12-19 paper/tsunami.tex
last changes from my side... i think

The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
\subsection{Parametric Uncertainty}
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the roughness parameter $n$. To this end,
we characterize the roughness $N$ by three different parameters
that are varied in three different regions but are assume
constant in each region. The three regions are as follows: 
on-shore who coefficnet is denoted by $N_1$, near-shore 
denoted by $N_2$, and deep-water denoted by $N_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $N$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $N \in [0.005-0.2]$.

>>>>

The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
\subsection{Parametric Uncertainty}
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the roughness parameter $n$. To this end,
we characterize the roughness $N$ by three different parameters
that are varied in three different regions but are assume
constant in each region. The three regions are as follows: 
on-shore who coefficnet is denoted by $N_1$, near-shore 
denoted by $N_2$, and deep-water denoted by $N_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $N$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $N \in [0.005-0.2]$.
